% Use of Generative AI Tools

\chapter{Use of Generative AI Tools}
\label{ch:genai}

As part of the research methodology, I used LLaMA 3.1 8B, LLaMA 3.3 70B, and Qwen 2.5 72B to generate synthetic survey responses. These models constitute the subject of the empirical investigation and are described in detail in Chapter~\ref{ch:method}.

In addition, I used the following generative AI tools for thesis preparation in accordance with the guidelines of the University of Mannheim:

\begin{itemize}
    \item \textbf{Claude (Anthropic):} Used for code generation in Python, formatting of LaTeX source code, generating descriptive comments for Python scripts, structuring the Git repository, and paraphrasing complex paragraphs.
    \item \textbf{Grammarly:} Used for proofreading and checking grammatical errors in the written text.
    \item \textbf{DeepL:} Used for English grammar and wording choice.
\end{itemize}

All AI-assisted content was reviewed, verified, and adapted by me. The intellectual contribution, research design, analysis, and conclusions presented in this thesis are entirely my own work.

\chapter{Declaration}
\label{ch:declaration}

I assert that this paper was written by me personally and that I was not assisted in any way by someone else. Furthermore, I assert that this paper or parts thereof have not been submitted elsewhere, neither by me nor by others. When I consulted print or electronic sources and publications to draw upon the writings or thoughts of others, I cited these sources. All secondary literature and additional sources have been acknowledged and are listed in the bibliography. The same is true for graphs, pictures, and all internet sources. Moreover, I consent that my paper may be screened and saved electronically and anonymously in order to be checked for plagiarism. I am aware that my paper may not be graded if I refuse to agree to these conditions.

\vspace{2cm}

\noindent
\begin{tabular}{@{}p{6cm}p{6cm}@{}}
\rule{5cm}{0.4pt} & \rule{5cm}{0.4pt} \\
Milana Taova & Place, Date
\end{tabular}
